---
title: 'Project 1: Chess Data'
author: "Aaron Grzasko"
date: "September 25, 2016"
output: 
    html_document:
        theme: cerulean
        highlight: pygments
---

---
## Assignment Overview  
In this assignment, we continue our work with semi-structured text data.  The goal of the exercise is to manipulate the provided *tournament.txt* document--a file containing chess tournament data--so that the following information is organized in a cleaner, *.csv* format:  
  
* player name
* player state
* total number of points scored in the chess tournament  
* player pre-tournament rating  
* average pre-chess rating of each player's opponents  
  
##  References  
The raw text file used in this assignment was originally retrieved from the course [Blackboard website](https://bbhosted.cuny.edu/bbcswebdav/pid-25842547-dt-content-rid-119781398_1/courses/SPS01_DATA_607_01_1169_1/SPS01_DATA_607_01_1169_1_ImportedContent_20160815114002/SPS01_DATA_607_01_1162_1_ImportedContent_20160126062155/tournamentinfo.txt).
  
The data is also available at my [Github Repo](https://raw.githubusercontent.com/spitakiss/Data607_Project1/master/tournamentinfo.txt)  

Additional assignment details are located [here](https://bbhosted.cuny.edu/bbcswebdav/pid-25842547-dt-content-rid-119781399_1/xid-119781399_1).

## Load Libraries  
  
The only package used in this exercise is the stringr package.

```{r load_lib}
library(stringr)
```
  
  
## Data Retrieval  

Retrieve the raw data directly from the specified Github repository:    
```{r data_retrieval, warning=FALSE}
myurl <- "https://raw.githubusercontent.com/spitakiss/Data607_Project1/master/tournamentinfo.txt"
raw_data <- readLines(myurl)
```

If the above code fails, we can manually save the .txt file to the working directory and run the following code:  
```{r alt_data_retrieval}
# uncomment code below if previous script fails
# raw_data <- readLines("tournamentinfo.txt")
```

## View Raw Data  

Using the short scripts below, we see that the raw data file contains 196 total lines.  
The data necessary for this assignment start on line 5 of the raw file.

Some variables of interest (e.g. player name, player total points ) begin on line 5, and repeat every three lines.  
Other variables (e.g. player state, pre-tournament points) begin on line 6, and again repeat on every third row.  
  
```{r raw_view, comment=NA}
head(raw_data,15)
length(raw_data)
``` 
  
## Data Manipulation  
  
The code chunk below stores the majority of the required variables into separate vectors, including:    
  
* player name
* total points scored
* player state
* player pre-rating  
  
The following code also creates multiple intermediate variables that will be useful for calculating the average, pre-rating of each player's opponents.  These variables include:  
  
* player number
* number of games played
* opponent list, which stores a vector of opponents separately for each player  
  
**Initial Variable Creation/Manipulation**  
```{r bulk_work}

# initialize variables in located in  lines 5, 8, 11, etc. of the raw file 

play_num <- vector()    # player number.  Intermediate varaible used to calculate opponent pre-scores
name <- vector()        # player name
ttl_pts <- vector()     # total points scored by player
game_num <- vector()    # number of games played (including wins, losses, and draws only).  
                        # this is an intermediate variable used for calculating average opponent pre-scores                             
opp_list <- list()      # opponent list, used to determine average opponent pre-scores

# initialize variables in lines 6,9, 12, etc. of raw file

state <- vector()       # player state of residence
pre_rating <- vector()  # player pre-tournament point ranking

for (i in 5:length(raw_data))
{
    if ((i - 2) %% 3 == 0)
    {
        # populate player number vector
        temp_play_num <- unlist(str_split(raw_data[i],"\\|"))[1]
        temp_play_num <- str_extract(temp_play_num,"[[0-9]]+")
        play_num <- c(play_num, temp_play_num)  
          
        # populate player name vector
        temp_name <- unlist(str_split(raw_data[i],"\\|"))
        name <- c(name, str_trim(temp_name[2]))  
        
        # populate ttl pts scored vector
        temp_ttl_pts <- unlist(str_split(raw_data[i],"\\|"))
        ttl_pts <- c(ttl_pts, as.numeric(temp_ttl_pts[3]))
        
        # populate opponent vector for each player, append to master, opponent list
        opp_vect <- vector()
        
        # individual player game counter.  Initialized value is zero  
        ivl_ct <- 0
        
        for (j in 4:10) # loop through seven rounds of potential opponents
        {
            # create opponent vector, loop through rounds 
            temp_opp <- unlist(str_split(raw_data[i],"\\|"))[j]
            opp_vect <- c(opp_vect, str_extract(temp_opp,"[[0-9]]+"))
            
            # calc final number of games, loop through rounds
            temp_game_num <- str_split(raw_data[i],"\\|")[[1]][j]
            ivl_ct <- str_detect(temp_game_num,"[WLD]") + ivl_ct
        }
        
        # final opponent list
        opp_list[[length(opp_list) + 1]] <- opp_vect
        
        # final game tally per player
        game_num <- c(game_num, ivl_ct)
    }
    else if (i %% 3 == 0)
    {
        # populate player state
        temp_state <- unlist(str_split(raw_data[i],"\\|"))
        state <- c(state, str_trim(temp_state[1]))   
        
        # populate player pre rating
        temp_pre_rating <- unlist(str_split(raw_data[i],"\\|"))[2]
        temp_pre_rating <- str_extract(temp_pre_rating,"R: [ [0-9]]+")  
        temp_pre_rating <- str_extract(temp_pre_rating,"[[0-9]]+")
        pre_rating <- c(pre_rating, str_trim(temp_pre_rating))
    }
}

```
<br>
**Data Checks**  

Let's do a brief visual check to confirm the data has been manipulated correctly.  
```{r}
head(play_num)
head(name)
head(state)
head(ttl_pts)
head(pre_rating)
head(opp_list,3)
head(game_num)
```
<br>  

Let's also confirm that there are 64 elements for each variable.  
```{r}
length(play_num)
length(name)
length(state)
length(ttl_pts)
length(pre_rating)
length(opp_list)
length(game_num)

```


**Create Data Frame**  
Next, we'll combine vectors into a data frame and create a variable to calculate each player's average opponent pre-score: 
<br>  
```{r create_df}
# create data frame
chess.df <- data.frame(play_num, name, state, ttl_pts = as.numeric(ttl_pts), pre_rating = as.numeric(pre_rating),
                       game_num = as.integer(game_num))

# for each player, sum up all opponenet pre-ratings, and store in vector, sum_opp_pre
sum_opp_pre <- vector()
for (i in 1:length(opp_list))
{
    temp_sum_rating <- sum(chess.df[which(chess.df$play_num %in% opp_list[[i]]),]$pre_rating)
    sum_opp_pre <- c(sum_opp_pre,temp_sum_rating)
}

# add new column to data frame, sum of opponent scores
chess.df$opp_sum_pre <- sum_opp_pre

# add another data frame column that calculates average opponent pre-score 
chess.df$avg_opp_pre <- round(chess.df$opp_sum_pre / chess.df$game_num,0)


```

**Write to CSV**  
Finally, we'll write a subset of the original data frame--containing only key variables--to csv format.  
  
```{r make_csv, comment=NA}

# subset the data frame to include only columns we need for exporting
chess.df.redux <- subset(chess.df, select = c(name,state,ttl_pts,pre_rating,avg_opp_pre))

# write to csv format
write.csv(chess.df.redux, row.names = FALSE)

# uncomment below to write file to working directory. 
# write.csv(chess.df.redux, file = "chess_data.csv", row.names=FALSE)
```


